### 2025-01-23 manuscript/manuscript.tex
update refs

@@

Examples of platforms that host open data soil data are the 
European Soil Data Centre \parencite{Panagos2022} and the World Soil Information
Service (WoSIS) of the ISRIC \parencite{Batjes2024}. Regarding the latter, a unified
platforms called SoilGrids provides open access to many predicted maps \parencite{poggio-soil-7-217-2021}.
Apart from samplings and literature,
spatial data are openly available terrestrial ecosystems.
To name a few, there openly available climatic, land cover, desertification risk, aridity, soil type, normalized
vegetation index, bedrock geological formations data.

@@

The samplings are distinguished as sequences, biodiversity and soil, using ENA, GBIF and WoSIS respectively.
The ENA API was used with a POST request containing the bounding box of Crete as a query \parencite{Yuan2023}. This 
query returned a list with the sample identifiers. Using this list and a custom python script the API
was invoked with GET requests to retrieve all the available metadata for each sample.
GBIF has incorporated in the online user interface, the functionality of selecting the area of interest with a bounding box and then requesting
the data \parencite{noauthor_gbif_nodate}. The request was first approved and then it was downloaded. 
IUCN redlist data were downloaded from the web interface \parencite{iucn2024}.
Advanced filters were selected, for Land Regions option \textit{Kriti} was opted 
along with global geographical scope. This search has a permalink on the user's profile.

@@

Plants have been studied in Crete since 300 BC \parencite{Krimbas_2005}. 
Extensive catalogues were compiled from Dioscorides during the 1st century AD. 
The Flora Graeca by John Sibthorp was published at 1806 and is widely 
considered one of most important publications about flora worldwide.
The Flora Graeca was innovative with great visualisations. 
Albeit currently, the botanic community appears to be behind 
in adopting open data standards. The most updated checklists 
are in hard-copy format 
\parencite{dimopoulos2016,strid2024atlas,ralph1993flora}.
Crete has more than 2200 native plants with 183 being Cretan endemics. 
The endemics have more than 8500 occurrences \parencite{kougioumoutzis2020plant}.
These monumental works is imperative that are uploaded to GBIF or similar
platforms to assist plant assessments and future analyses.

Worldwide projects of microbiome studies have collected one or two topsoil
samples from Crete \parencite{Vasar2022, Labouyrie2023, Bahram2018, Orgiazzi2018}.
Some have focused on soil fungi \parencite{Mikryukov2023, Davison2021, Tedersoo2021}
and other soil eukaryotes \parencite{Aslani2022}.
The only thorough microbiome study of a soil ecosystem in Crete, to our knowledge,
is in the north west part of the island, the Koiliaris Critical Zone Observatory \parencite{tsiknia2014}.

@@

Apart from the soil microbiome, soil physical and chemical properties has been
investigated by global and European projects like the Forum of European Geological Surveys
(FOREGS) \parencite{nerc19017}, the Geochemical Mapping of Agricultural and Grazing Land
Soil in Europe (GEMAS) \parencite{REIMANN2018302} and the Soil Profile Analytical
Database for Europe (SPADE) \parencite{Hiederer2006}. These data with other data
not publicly available have led to the creation of the \href{https://esdac.jrc.ec.europa.eu/content/soil-map-greece-0}{soil map} of Crete 
and Greece in general. These maps are not as complete as the geological maps (Figure \ref{fig:crete_data_cube_map} C), 
and that is the reason of the coarse mapping of the Harmonized World Soil Database map of Crete (Figure \ref{fig:crete_data_cube_map} D).
Soil type results highlight the diversity of soil types on Crete, with a
dominance of Lithic Leptosols and Calcaric Regosols, which may be linked to the
region's geomorphological features (karst, limestone).

>>>>

edaphobase 2 \parencite{russell2024edaphobase}

assessing disparites \parencite{ellis-soto2024addressing}

soil geography history \parencite{rodrigo-comino2018the-multidisciplinary}

@@

Examples of platforms that host open data soil data are the 
European Soil Data Centre \parencite{panagos2022european} and the World Soil Information
Service (WoSIS) of the ISRIC \parencite{Batjes2024}. Regarding the latter, a unified
platforms called SoilGrids provides open access to many predicted maps \parencite{poggio-soil-7-217-2021}.
Apart from samplings and literature,
spatial data are openly available terrestrial ecosystems.
To name a few, there openly available climatic, land cover, desertification risk, aridity, soil type, normalized
vegetation index, bedrock geological formations data.

@@

The samplings are distinguished as sequences, biodiversity and soil, using ENA, GBIF and WoSIS respectively.
The ENA API was used with a POST request containing the bounding box of Crete as a query \parencite{yuan2023the-european}. This 
query returned a list with the sample identifiers. Using this list and a custom python script the API
was invoked with GET requests to retrieve all the available metadata for each sample.
GBIF has incorporated in the online user interface, the functionality of selecting the area of interest with a bounding box and then requesting
the data \parencite{noauthor_gbif_nodate}. The request was first approved and then it was downloaded. 
IUCN redlist data were downloaded from the web interface \parencite{iucn2024}.
Advanced filters were selected, for Land Regions option \textit{Kriti} was opted 
along with global geographical scope. This search has a permalink on the user's profile.

@@

Plants have been studied in Crete since 300 BC \parencite{krimbas2005h.w.}. 
Extensive catalogues were compiled from Dioscorides during the 1st century AD. 
The Flora Graeca by John Sibthorp was published at 1806 and is widely 
considered one of most important publications about flora worldwide.
The Flora Graeca was innovative with great visualisations. 
Albeit currently, the botanic community appears to be behind 
in adopting open data standards. The most updated checklists 
are in hard-copy format 
\parencite{dimopoulos2016,strid2024atlas,ralph1993flora}.
Crete has more than 2200 native plants with 183 being Cretan endemics. 
The endemics have more than 8500 occurrences \parencite{kougioumoutzis2020plant}.
These monumental works is imperative that are uploaded to GBIF or similar
platforms to assist plant assessments and future analyses.

Worldwide projects of microbiome studies have collected one or two topsoil
samples from Crete \parencite{Vasar2022, Labouyrie2023, Bahram2018, Orgiazzi2018}.
Some have focused on soil fungi \parencite{Mikryukov2023, Davison2021, Tedersoo2021}
and other soil eukaryotes \parencite{Aslani2022}.
The only thorough microbiome study of a soil ecosystem in Crete, to our knowledge,
is in the north west part of the island, the Koiliaris Critical Zone Observatory \parencite{tsiknia2014environmental}.

@@

Apart from the soil microbiome, soil physical and chemical properties has been
investigated by global and European projects like the Forum of European Geological Surveys
(FOREGS) \parencite{nerc19017}, the Geochemical Mapping of Agricultural and Grazing Land
Soil in Europe (GEMAS) \parencite{reimann2018gemas} and the Soil Profile Analytical
Database for Europe (SPADE) \parencite{Hiederer2006}. These data with other data
not publicly available have led to the creation of the \href{https://esdac.jrc.ec.europa.eu/content/soil-map-greece-0}{soil map} of Crete 
and Greece in general. These maps are not as complete as the geological maps (Figure \ref{fig:crete_data_cube_map} C), 
and that is the reason of the coarse mapping of the Harmonized World Soil Database map of Crete (Figure \ref{fig:crete_data_cube_map} D).
Soil type results highlight the diversity of soil types on Crete, with a
dominance of Lithic Leptosols and Calcaric Regosols, which may be linked to the
region's geomorphological features (karst, limestone).
